Noncoherent least squares estimators of carrier phase and amplitude

### 2012-10-16 other/confnoncoherent/paper.tex
fixed silly abs error
fixed abs errors

We will have particular use of complex random variables of the form $1 + X$ where $X$ is circularly symmetric.  Let $R \geq 0$ and $\Phi \in [0,2\pi)$ be real random variables satisfying, 
\[
R e^{j\Phi} = 1 + X.
\]
The joint distribution of $R$ and $\Phi$ can be shown to be
\[
f(r,\phi) = \frac{r f_Z(\sqrt{\sabs{r^2 - 2r\cos\phi + 1}})}{2\pi\sqrt{\sabs{r^2 - 2r\cos\phi + 1}}}.
\]
% The mean of $R e^{j\Phi}$ is equal to one because the mean of $X$ is zero.  So,
% \[
% \expect \Re(R e^{j\Phi}) = \expect R \cos(\Phi) = 1,
% \]
% where $\Re(\cdot)$ denotes the real part, and
% \[
% \expect \Im(R e^{j\Phi}) = \expect R \sin(\Phi) = 0,
% \]
% where $\Im(\cdot)$ denotes the imaginary part.  %The next lemma will be useful for the analysis of the least squares estimator.

@@

\begin{theorem}\label{thm:consistency} (Almost sure convergence)
Let $\{w_i\}$ be a sequence of independent and identically distributed, circularly symmetric complex random variables with $\expect \abs{w_1}^2$ finite, and let $y_1,\dots, y_L$ be given by~\eqref{eq:sigmod}.   Let $\hat{a} = \hat{\rho}e^{j\hat{\theta}}$ be the least squares estimator of $a_0 = \rho_0e^{j\theta_0}$.  Let $R_i \geq 0$ and $\Phi_i \in [0,2\pi)$ be real random variables satisfying
\begin{equation}\label{eq:RiandPhii}
R_ie^{j\Phi_i} = 1 + \frac{w_i}{a_0 s_i},
\end{equation}
and define the continuous function
\[
G(x) = \expect R_1 \cos\sfracpart{ x + \Phi_1}.
\] 
If $\abs{G(x)}$ is uniquely maximised at $x = 0$ over the interval $[-\tfrac{\pi}{M},\tfrac{\pi}{M})$, then:
\begin{enumerate}
\item $\sfracpart{\theta_0 - \hat{\theta}} \rightarrow 0$ almost surely as $L \rightarrow \infty$,
\item $\hat{\rho} \rightarrow \rho_0 G(0)$ almost surely as $L \rightarrow \infty$.
\end{enumerate}
\end{theorem}

@@

A key assumption in Theorem~\ref{thm:consistency} is that $\abs{G(x)}$ is uniquely maximised at $x = 0$.  The next proposition describes a class of circularly symmetric distributions for which $\abs{G(x)}$ is uniquely maximised at $x = 0$.  The zero mean complex Gaussian distribution with independent real and imaginary parts is in this class.  We omit the proof due to space constraints.

\begin{proposition}\label{prop:contgg2}
Let $W = X + jY$ be a circularly symmetric complex random variable, the real and imaginary parts having pdf $f_{X,Y}(x,y)$ that is nonincreasing with $x^2 + y^2$.  Let $R \geq 0$ and $\Phi \in [0, 2\pi)$ be random variables satisfying $R e^{j\Phi} = 1 + W$.  Let,
\[
G(x) =  \expect R \cos\sfracpart{x + \Phi}.
\]
Then $\abs{G(x)}$ is uniquely maximised at $x=0$ over the interval $[-\tfrac{\pi}{M},\tfrac{\pi}{M}]$.
\end{proposition}

@@

Figures~\ref{fig:plotphaseM2}~and~\ref{fig:plotphaseM4} show the mean square error (MSE) of the phase estimator $\hat{\theta}$ computed as $\tfrac{1}{T}\sum_{i=1}^T\sfracpart{\hat{\theta}_i - \theta_0}^2$.  The dots show the simulated MSE of the least square estimator.  The dashed line is the simulated MSE of the unmodulated carrier estimator~\eqref{eq:hataumod} that has a priori knowledge of the transmitted symbols $s_1, \dots, s_L$.  %The dashed line acts as a lower bound on performance.  
The solid line is the MSE predicted by Theorem~\ref{thm:normality}.  The theorem accurately predicts the behaviour of the phase estimator when $L$ is sufficiently large and when the SNR is not too small.  As the SNR decreases the variance of the phase estimator approaches that of the uniform distribution on $[-\tfrac{\pi}{M}, \tfrac{\pi}{M})$, and Theorem~\ref{thm:normality} does not model this behaviour.  %An interesting feature is that the phase estimator is not accurate at low SNR unless $L$ is very large.

Figures~\ref{fig:plotphaseM2}~and~\ref{fig:plotphaseM4} also display the simulated MSE of the phase estimator of Viterbi and Viterbi~\cite{ViterbiViterbi_phase_est_1983}.  This estimator requires the selection of a function $F : \reals \mapsto \reals$.  We have chosen $F(x) = 1$.  The MSE of the least squares estimator and the Viterbi and Viterbi estimator is similar.  The least squares estimator appears slightly more accurate at low SNR.

>>>>

We will have particular use of complex random variables of the form $1 + X$ where $X$ is circularly symmetric.  Let $R \geq 0$ and $\Phi \in [0,2\pi)$ be real random variables satisfying, 
\[
R e^{j\Phi} = 1 + X.
\]
The joint distribution of $R$ and $\Phi$ can be shown to be
\[
f(r,\phi) = \frac{r f_Z(\sqrt{r^2 - 2r\cos\phi + 1})}{2\pi\sqrt{r^2 - 2r\cos\phi + 1}}.
\]
% The mean of $R e^{j\Phi}$ is equal to one because the mean of $X$ is zero.  So,
% \[
% \expect \Re(R e^{j\Phi}) = \expect R \cos(\Phi) = 1,
% \]
% where $\Re(\cdot)$ denotes the real part, and
% \[
% \expect \Im(R e^{j\Phi}) = \expect R \sin(\Phi) = 0,
% \]
% where $\Im(\cdot)$ denotes the imaginary part.  %The next lemma will be useful for the analysis of the least squares estimator.

@@

\begin{theorem}\label{thm:consistency} (Almost sure convergence)
Let $\{w_i\}$ be a sequence of independent and identically distributed, circularly symmetric complex random variables with $w_1$ having finite variance and continuous pdf.  Let $y_1,\dots, y_L$ be given by~\eqref{eq:sigmod} and let $\hat{a} = \hat{\rho}e^{j\hat{\theta}}$ be the least squares estimator of $a_0 = \rho_0e^{j\theta_0}$.  Let $R_i \geq 0$ and $\Phi_i \in [0,2\pi)$ be real random variables satisfying
\begin{equation}\label{eq:RiandPhii}
R_ie^{j\Phi_i} = 1 + \frac{w_i}{a_0 s_i},
\end{equation}
and define the continuous function
\[
G(x) = \expect R_1 \cos\sfracpart{ x + \Phi_1}.
\] 
If $G(x)$ is uniquely maximised at $x = 0$ over the interval $[-\tfrac{\pi}{M},\tfrac{\pi}{M})$, then:
\begin{enumerate}
\item $\sfracpart{\theta_0 - \hat{\theta}} \rightarrow 0$ almost surely as $L \rightarrow \infty$,
\item $\hat{\rho} \rightarrow \rho_0 G(0)$ almost surely as $L \rightarrow \infty$.
\end{enumerate}
\end{theorem}

@@

A key assumption in Theorem~\ref{thm:consistency} is that $G(x)$ is uniquely maximised at $x = 0$.  The next proposition describes a class of circularly symmetric distributions for which $G(x)$ is uniquely maximised at $x = 0$.  The zero mean complex Gaussian distribution with independent real and imaginary parts is in this class.  We omit the proof due to space constraints.

\begin{proposition}\label{prop:contgg2}
Let $W = X + jY$ be a circularly symmetric complex random variable, the real and imaginary parts having pdf $f_{X,Y}(x,y)$ that is nonincreasing with $x^2 + y^2$.  Let $R \geq 0$ and $\Phi \in [0, 2\pi)$ be random variables satisfying $R e^{j\Phi} = 1 + W$.  Let,
\[
G(x) =  \expect R \cos\sfracpart{x + \Phi}.
\]
Then $G(x)$ is uniquely maximised at $x=0$ over the interval $[-\tfrac{\pi}{M},\tfrac{\pi}{M}]$.
\end{proposition}

@@

Figures~\ref{fig:plotphaseM2}~and~\ref{fig:plotphaseM4} show the sample mean square error (MSE) of the phase estimator $\hat{\theta}$ computed as $\tfrac{1}{T}\sum_{i=1}^T\sfracpart{\hat{\theta}_i - \theta_0}^2$.  The dots show the sample MSE of the least square estimator.  The dashed line is the sample MSE of the unmodulated carrier estimator~\eqref{eq:hataumod} that has a priori knowledge of the transmitted symbols $s_1, \dots, s_L$.  %The dashed line acts as a lower bound on performance.  
The solid line is the MSE predicted by Theorem~\ref{thm:normality}.  The theorem accurately predicts the behaviour of the phase estimator when $L$ is sufficiently large and when the SNR is not too small.  As the SNR decreases the variance of the phase estimator approaches that of the uniform distribution on $[-\tfrac{\pi}{M}, \tfrac{\pi}{M})$, and Theorem~\ref{thm:normality} does not model this behaviour.  %An interesting feature is that the phase estimator is not accurate at low SNR unless $L$ is very large.

Figures~\ref{fig:plotphaseM2}~and~\ref{fig:plotphaseM4} also display the sample MSE of the phase estimator of Viterbi and Viterbi~\cite{ViterbiViterbi_phase_est_1983}.  This estimator requires the selection of a function $F : \reals \mapsto \reals$.  We have chosen $F(x) = 1$.  The sample MSE of the least squares estimator and the Viterbi and Viterbi estimator is similar.  The least squares estimator appears slightly more accurate at low SNR.
